PanGraph: scalable bacterial pan-genome graph construction

### 2022-10-03 main.tex
completed draft of marginalize paragraph

Interpretation of a pangenome graph containing hundreds of strains can be challenging. To this end it is often informative to inspect simpler sub-graphs containing only a small subset of isolates. However building multiple such sub-graphs can be computationally intensive. To facilitate this task \textit{PanGraph} provides the \verb|marginalize| command, that can be used to marginalize a large pangenome graph on a small subset of strains, removing all the other paths and merging transitive pancontigs. This operation is computationally much cheaper than building a new graph for the subset of strains considered.\\
To verify that marginalized graphs are compatible with newly-built graphs, we built a pangenome graph for 50 randomly selected strains from the \textit{Klebsiella Pneumoniae} dataset. We then picked 50 random pairs of strains from the same set and built two graphs for each pair: the graph obtained by marginalizing the complete graph on the pair, and the graph built directly from the pair (cf. Fig.~\ref{fig:marginalization}A, top).
We then compared these two graphs by considering the partition they generate on the genome of the isolates they include. Each genome is partitioned in pancontigs that are either shared on the pair, or private to one isolate
(cf. Fig.~\ref{fig:marginalization}A, bottom). We can classify the segments in the intersection of the two partitions in two categories: either the two partitions agree or they do not on the segment kind. And agreement can be further divided in agreement on the segments being shared or agreement on them being private.
% TODO: Compatibility between the two graphs would required segments on which the two graphs disagree to be short and few.\\

% TODO: cite supplementary

\begin{figure*}[htb]
    \includegraphics[width=\textwidth]{figs/marginalize_test.pdf}
    \caption{{\bf Test of graph marginalization}.}
    \label{fig:marginalization}
\end{figure*}

>>>>

The interpretation of a pangenome graph containing hundreds of strains can be challenging. To this end it is often informative to inspect simpler sub-graphs comprising only a small subset of isolates. However building multiple such sub-graphs is computationally intensive. To facilitate this task \textit{PanGraph} provides the \verb|marginalize| command, that can be used to marginalize a large pangenome graph on a small subset of strains, removing all the other paths and merging transitive pancontigs. This operation is computationally much cheaper than building a new graph for the subset of strains considered.\\
To verify that marginalized graphs are compatible with newly-built graphs, we built a pangenome graph for 50 randomly selected strains from the \textit{Klebsiella Pneumoniae} dataset. We then picked 50 random pairs of strains from the same set and built two graphs for each pair: the graph obtained by marginalizing the complete graph on the pair, and the graph built directly from the pair (cf. Fig.~\ref{fig:marginalization}A, top).
We compare these two graphs by considering the partition they generate on the genome of the isolates they include. Each genome is partitioned in pancontigs that are either shared on the pair, or private to one isolate (cf. Fig.~\ref{fig:marginalization}A, bottom). We can classify the segments in the intersection of the two partitions in different categories, depending on whether the two partitions agree on whether the segment is shared or private. Moreover segments on which the two partitions agree are further sub-divided depending on whether they are private or shared on both partitions.\\
Compatibility between the two graphs requires segments on which the two partitions disagree to be few and short. Indeed, we verified that over the pairs we picked these segments cover a very small fraction of the genome ($<1\%$), and have average size compatible with the default 100 bp precision threshold of \textit{PanGraph} (cf. Fig.~\ref{fig:marginalization}B-C). This is contrary to agreement segments, that have average size of several kbps. The prevalence of agreement segment cannot be explained only by the fact that most of the genome is shared, since segments that are shared on both graphs cover on average only 85\% of the sequence. We confirmed this results using the fraction of shared k-mers, using an approach inspired by PopPUNK \cite{lees2019fast}. Namely, we approximate the fraction of homologous sequence between any two pairs using the fraction of shared k-mers (in our case we take $k=21$). This fraction is however an underestimate of the fraction of homologous sequence because of mutations. We therefore correct it with the factor $(1-d)^{-k}$, where $d$ is the average pairwise divergence on core genes for the pair considered. The resulting distribution of shared sequence is indeed in very good agreement with the fraction of segments that are shared on both graphs (cf. Fig.~\ref{fig:marginalization}B), suggesting that homologous sequences are correctly merged on the complete, marginalized and pairwise graphs.\\
We performed the same test on species from the other datasets, obtaining similar results on all species whose divergence is compatible with \textit{PanGraph} capabilities (cf. SI Section IV and SI Fig.~5).

\begin{figure*}[htb]
    \includegraphics[width=\textwidth]{figs/marginalize_test.pdf}
    \caption{{\bf Test of graph marginalization}.
        \textbf{A}: we built a pangenome graph from 50 randomly chosen strains from the \textit{K. Pneumoniae} dataset. We then randomly picked 50 pairs of strains. For each pair we compared the pangenome graph obtained by marginalizing the complete graph on the pair of strain, and the one obtained by building a new graph for the pair (top). The comparison is done by considering that each graph partitions a genome in shared and private segments. By combining the partitions generated by the marginalized and pairwise graphs we categorize segments in three categories, depending on whether the two partitions agree or not, and if they agree depending on whether segments are shared or private. All graphs were built using \textit{minimap2} alignment kernel with \textit{asm20} option and default value for the energy parameters.
        \textbf{B}: distribution of the average fraction of the genome covered by segments of each category, over the 50 pairs considered (2 entries per pair). The last line represents the distribution of shared sequence, approximated using the fraction of shared k-mers corrected using sequence divergence as described in the main text.
        \textbf{C}: distribution of average segment lengths for each category over the 50 pairs considered (2 entries per pair).
    }
    \label{fig:marginalization}
\end{figure*}
